---
title: "GSD and GP regression - week 6"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
date: "2025-11-13"
---

```{css, echo=FALSE}
#TOC {
    max-width: fit-content;
    white-space: nowrap;
}
  
div:has(> #TOC) {
    display: flex;
    flex-direction: row-reverse;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load required libraries
library(mvtnorm)
library(ggplot2)

# set the seed
set.seed(585883357)
```

# Load the GSD package

```{r}
devtools::load_all(path = "../groupSequentialDesigns/")
```

Contains:

1. `calc_bounds_of()`
2. `calc_bounds_pocock()`
3. `calc_bounds_triangular()`
4. `gsd_simulations()`

# Calculate staged power

Calculate the boundaries.

```{r}
pocock_bounds <- calc_bounds_pocock()
of_bounds <- calc_bounds_of()
tri_bounds <- calc_bounds_triangular()
```

Get values for ESS, VSS, and power.

First, generate $\delta$ vector.

```{r}
deltas <- seq(from = 0, to = 1, by = 0.001)
```

```{r}
gsd_simulations(upper_bounds = pocock_bounds$upper_bounds,
                         lower_bounds = pocock_bounds$lower_bounds,
                         alt_hypothesis = deltas[1],
                         n_patients = c(19.51241 * 1:3))$analysis_1["efficacy_alt"]
```


## Pocock

```{r}
pk_analysis_1 <- vector(mode = "numeric", length = length(deltas))
pk_analysis_2 <- vector(mode = "numeric", length = length(deltas))
pk_analysis_3 <- vector(mode = "numeric", length = length(deltas))

for (i in 1:length(deltas)) {
  
  sim <- gsd_simulations(upper_bounds = pocock_bounds$upper_bounds,
                         lower_bounds = pocock_bounds$lower_bounds,
                         alt_hypothesis = deltas[i],
                         n_patients = c(19.51241 * 1:3))
  
  pk_analysis_1[i] <- sim$analysis_1["efficacy_alt"]
  pk_analysis_2[i] <- sim$analysis_2["efficacy_alt"]
  pk_analysis_3[i] <- sim$analysis_3["efficacy_alt"]
}
```


## O'Brien-Fleming

```{r}
of_analysis_1 <- vector(mode = "numeric", length = length(deltas))
of_analysis_2 <- vector(mode = "numeric", length = length(deltas))
of_analysis_3 <- vector(mode = "numeric", length = length(deltas))

for (i in 1:length(deltas)) {
  
  sim <- gsd_simulations(upper_bounds = of_bounds$upper_bounds,
                         lower_bounds = of_bounds$lower_bounds,
                         alt_hypothesis = deltas[i],
                         n_patients = c(16.93269 * 1:3))
  
  of_analysis_1[i] <- sim$analysis_1["efficacy_alt"]
  of_analysis_2[i] <- sim$analysis_2["efficacy_alt"]
  of_analysis_3[i] <- sim$analysis_3["efficacy_alt"]
}
```


## Triangular

```{r}
tri_analysis_1 <- vector(mode = "numeric", length = length(deltas))
tri_analysis_2 <- vector(mode = "numeric", length = length(deltas))
tri_analysis_3 <- vector(mode = "numeric", length = length(deltas))

for (i in 1:length(deltas)) {
  
  sim <- gsd_simulations(upper_bounds = tri_bounds$upper_bounds,
                         lower_bounds = tri_bounds$lower_bounds,
                         alt_hypothesis = deltas[i],
                         n_patients = c(20.43845 * 1:3))
  
  tri_analysis_1[i] <- sim$analysis_1["efficacy_alt"]
  tri_analysis_2[i] <- sim$analysis_2["efficacy_alt"]
  tri_analysis_3[i] <- sim$analysis_3["efficacy_alt"]
}
```

All of the above have a cumulative power at the end of the trial of approximately 80%. At each stage, power is cumulative sum of the probability of claiming efficacy under the alternative.

## Plots

### Stage 1

```{r}
ggplot() +
  geom_line(mapping = aes(x = deltas, y = pk_analysis_1, color = "Pocock")) + 
  geom_line(mapping = aes(x = deltas, y = of_analysis_1, color = "O'Brien-Fleming")) + 
  geom_line(mapping = aes(x = deltas, y = tri_analysis_1, color = "Triangular")) +
  ylim(c(0,1)) +
  scale_color_manual(name = "Boundary types",
                     breaks = c("Pocock", "O'Brien-Fleming", "Triangular"),
                     values = c("Pocock" = "red",
                                "O'Brien-Fleming" = "blue",
                                "Triangular" = "purple")) +
  labs(x = "Alternative hypothesis = \u03B4", 
       y = "Power at first analysis", 
       title = "Power for different boundaries at Stage 1")
```

### Stage 2

```{r}
ggplot() +
  geom_line(mapping = aes(x = deltas, y = pk_analysis_2, color = "Pocock")) + 
  geom_line(mapping = aes(x = deltas, y = of_analysis_2, color = "O'Brien-Fleming")) + 
  geom_line(mapping = aes(x = deltas, y = tri_analysis_2, color = "Triangular")) +
  ylim(c(0,1)) +
  scale_color_manual(name = "Boundary types",
                     breaks = c("Pocock", "O'Brien-Fleming", "Triangular"),
                     values = c("Pocock" = "red",
                                "O'Brien-Fleming" = "blue",
                                "Triangular" = "purple")) +
  labs(x = "Alternative hypothesis = \u03B4", 
       y = "Power at second analysis", 
       title = "Power for different boundaries at Stage 2")
```

### Stage 3

```{r}
ggplot() +
  geom_line(mapping = aes(x = deltas, y = pk_analysis_3, color = "Pocock")) + 
  geom_line(mapping = aes(x = deltas, y = of_analysis_3, color = "O'Brien-Fleming")) + 
  geom_line(mapping = aes(x = deltas, y = tri_analysis_3, color = "Triangular")) +
  ylim(c(0,1)) +
  scale_color_manual(name = "Boundary types",
                     breaks = c("Pocock", "O'Brien-Fleming", "Triangular"),
                     values = c("Pocock" = "red",
                                "O'Brien-Fleming" = "blue",
                                "Triangular" = "purple")) +
  labs(x = "Alternative hypothesis = \u03B4", 
       y = "Power at third analysis", 
       title = "Power for different boundaries at Stage 3")
```


## Cumulative sum plots

### Stage 1

```{r}
ggplot() +
  geom_line(mapping = aes(x = deltas, y = pk_analysis_1, color = "Pocock")) + 
  geom_line(mapping = aes(x = deltas, y = of_analysis_1, color = "O'Brien-Fleming")) + 
  geom_line(mapping = aes(x = deltas, y = tri_analysis_1, color = "Triangular")) +
  ylim(c(0,1)) +
  scale_color_manual(name = "Boundary types",
                     breaks = c("Pocock", "O'Brien-Fleming", "Triangular"),
                     values = c("Pocock" = "red",
                                "O'Brien-Fleming" = "blue",
                                "Triangular" = "purple")) +
  labs(x = "Alternative hypothesis = \u03B4", 
       y = "Power at first analysis", 
       title = "Power for different boundaries at Stage 1")
```

### Stage 2

```{r}
ggplot() +
  geom_line(mapping = aes(x = deltas, y = pk_analysis_1+pk_analysis_2, color = "Pocock")) + 
  geom_line(mapping = aes(x = deltas, y = of_analysis_1+of_analysis_2, color = "O'Brien-Fleming")) + 
  geom_line(mapping = aes(x = deltas, y = tri_analysis_1+tri_analysis_2, color = "Triangular")) +
  ylim(c(0,1)) +
  scale_color_manual(name = "Boundary types",
                     breaks = c("Pocock", "O'Brien-Fleming", "Triangular"),
                     values = c("Pocock" = "red",
                                "O'Brien-Fleming" = "blue",
                                "Triangular" = "purple")) +
  labs(x = "Alternative hypothesis = \u03B4", 
       y = "Power at second analysis", 
       title = "Power for different boundaries at Stage 2")
```

### Stage 3

```{r}
ggplot() +
  geom_line(mapping = aes(x = deltas, y = pk_analysis_1+pk_analysis_2+pk_analysis_3, color = "Pocock")) + 
  geom_line(mapping = aes(x = deltas, y = of_analysis_1+of_analysis_2+of_analysis_3, color = "O'Brien-Fleming")) + 
  geom_line(mapping = aes(x = deltas, y = tri_analysis_1+tri_analysis_2+tri_analysis_3, color = "Triangular")) +
  ylim(c(0,1)) +
  scale_color_manual(name = "Boundary types",
                     breaks = c("Pocock", "O'Brien-Fleming", "Triangular"),
                     values = c("Pocock" = "red",
                                "O'Brien-Fleming" = "blue",
                                "Triangular" = "purple")) +
  labs(x = "Alternative hypothesis = \u03B4", 
       y = "Power at third analysis", 
       title = "Power for different boundaries at Stage 3")
```


# David's Triangular test code

## 1. Define the Boundary Function

```{r}
get_boundaries <- function(alpha, delta, Imax, K) {
  # Information levels at each look k=1...K
  k <- 1:K
  I_k <- (k / K) * Imax
  
  # The correction factor for discrete monitoring (0.583 is the specific constant)
  correction <- 0.583 * sqrt(Imax / K)
  
  # Common intercept term (A)
  intercept <- (2 / delta) * log(1 / (2 * alpha))
  
  # Calculate Upper Boundary (dk)
  # Formula: (2/delta)*log(1/(2*alpha)) - 0.583*sqrt(Imax/K) + (delta/4)*I_k
  dk <- intercept - correction + (delta / 4) * I_k
  
  # Calculate Lower Boundary (ck)
  # Formula: -(2/delta)*log(1/(2*alpha)) + 0.583*sqrt(Imax/K) + (3*delta/4)*I_k
  ck <- -intercept + correction + (3 * delta / 4) * I_k
  
  return(list(k = k, I_k = I_k, lower = ck, upper = dk))
}
```


## 2. Simulation Function

```{r}
simulate_gst <- function(n_sim, alpha, delta, Imax, K, true_theta) {
  
  # Get boundaries
  bounds <- get_boundaries(alpha, delta, Imax, K)
  upper <- bounds$upper
  lower <- bounds$lower
  
  # Information increment per stage
  info_increment <- Imax / K
  
  # Initialize counters
  rejected <- 0
  accepted <- 0 # Accepted Null (futility)
  overrun  <- 0 # Reached max K without crossing (should be rare in Triangular designs)
  
  # Standard Error of the increment (sqrt of information increment)
  sd_increment <- sqrt(info_increment)
  # Mean of the increment (theta * information increment)
  mean_increment <- true_theta * info_increment
  
  # Run Simulation (Vectorized loop for efficiency)
  # We generate the full paths, then check stopping conditions
  
  # Generate all increments: matrix of size (n_sim x K)
  increments <- matrix(rnorm(n_sim * K, mean = mean_increment, sd = sd_increment), 
                       nrow = n_sim, ncol = K)
  
  # Calculate cumulative score statistic Sk
  # t(apply(...)) is used because apply returns column-vectors
  Sk <- t(apply(increments, 1, cumsum)) 
  
  # Evaluate stopping criteria
  for (i in 1:n_sim) {
    path <- Sk[i, ]
    stopped <- FALSE
    
    for (stage in 1:K) {
      if (path[stage] >= upper[stage]) {
        rejected <- rejected + 1
        stopped <- TRUE
        break
      } else if (path[stage] <= lower[stage]) {
        accepted <- accepted + 1
        stopped <- TRUE
        break
      }
    }
    
    # Check if trial finished without crossing boundaries
    if (!stopped) {
      # In a closed triangular test, usually the boundaries meet.
      # If they haven't met by K and we haven't crossed, we need a convention.
      # Convention: If Sk > (upper+lower)/2, Reject. Otherwise Accept.
      # For this simulation, we simply track it as 'overrun' or failure to reject.
      overrun <- overrun + 1
    }
  }
  
  return(list(
    rate = rejected / n_sim,
    overrun_rate = overrun / n_sim,
    n_sim = n_sim
  ))
}
```


## 3. Execution and Visualization

```{r}
# Parameters
set.seed(123)

alpha_param <- alpha <- 0.05
delta_param <- 0.693147   # Effect size (e.g., log-hazard ratio or mean diff)

K <- 3

beta <- 0.3

delta_param <- 2*qnorm(1-alpha)*delta_param/(qnorm(1-alpha)+qnorm(1-beta))


Imax_param  <- (sqrt(4*0.583^2/K + 8*log(1/(2*alpha))) - 2*0.583/sqrt(K))^2/delta_param^2    # Max Information (depends on variance and N)

K_param     <- 3   # Number of looks
N_SIM       <- 10000

# Get Boundaries for plotting
b <- get_boundaries(alpha_param, delta_param, Imax_param, K_param)

# --- Run Simulation: Power (True Theta = Delta) ---
res_power <- simulate_gst(N_SIM, alpha_param, delta_param, Imax_param, K_param, 
                          true_theta = 0.693147)

# --- Run Simulation: Type I Error (True Theta = 0) ---
res_type1 <- simulate_gst(N_SIM, alpha_param, delta_param, Imax_param, K_param, 
                          true_theta = 0)

# 4. Print Results
cat("==========================================\n")
cat("Group Sequential Trial Simulation Results\n")
cat("==========================================\n")
cat(sprintf("Parameters: alpha=%.3f, delta=%.2f, Imax=%.1f, K=%d\n", 
            alpha_param, delta_param, Imax_param, K_param))
cat("------------------------------------------\n")
cat(sprintf("Estimated Power (Theta = %.2f):     %.4f\n", delta_param, res_power$rate))
cat(sprintf("Estimated Type I Error (Theta = 0): %.4f\n", res_type1$rate))
cat("==========================================\n")
b
```


```{r}
dave_test <- calc_bounds_triangular(n_analyses = 5, alpha = 0.05, delta = 0.66)
dave_test
```

```{r}
b$upper
b$lower
```


```{r}
gsd_simulations(n_analyses = 3, upper_bounds = b$upper, lower_bounds = b$lower,
                n_patients = c(100, 200, 300))
```


# Mazin's New Triangular bounds

```{r}
calc_bounds_triangular_new <- function(n_analyses = 3,
                                       alpha = 0.05,
                                       power = 0.7,
                                       delta = 0.693147) {
  
  beta <- 1-power
  
  delta_tilde <- (2 * qnorm(1-alpha) * delta) / (qnorm(1-alpha) + qnorm(1-beta))
  
  I_L_term1 <- (4 * (0.583^2)) / n_analyses
  I_L_term2 <- 8 * log(1/(2*alpha))
  I_L_term3 <- (2 * 0.583) / sqrt(n_analyses)
  
  # replace delta with delta_tilde
  I_L <- (sqrt(I_L_term1 + I_L_term2) - I_L_term3)^2 * (1 / delta_tilde)^2
  
  # replace delta with delta_tilde
  bounds_term1 <- (2/delta_tilde) * log(1/(2*alpha))
  
  # unchanged
  bounds_term2 <- 0.583 * sqrt(I_L / n_analyses)
  
  analysis_fracs <- (1:n_analyses / n_analyses)
  I_L_fracs <- I_L * analysis_fracs
  
  # replace delta with delta_tilde
  e_l <- (bounds_term1 - bounds_term2 + ((0.25*delta_tilde) * analysis_fracs * I_L )) / sqrt(I_L_fracs)
  
  # replace delta with delta_tilde
  f_l <- (-bounds_term1 + bounds_term2 + ((0.75*delta_tilde) * analysis_fracs * I_L )) / sqrt(I_L_fracs)
  
  # return delta_tilde as well
  return(list(upper_bounds = e_l, 
              lower_bounds = f_l, 
              delta_tilde = delta_tilde,
              info = I_L_fracs))
  
  
}
```


```{r}
calc_bounds_triangular_new(power=0.9)
```


```{r}
gsd_simulations(n_analyses = 3, upper_bounds = calc_bounds_triangular_new()$upper_bounds,
                lower_bounds = calc_bounds_triangular_new()$lower_bounds,
                n_patients = c(10,20,30), null_hypothesis = 0, alt_hypothesis = 0.7791986)
```


The difference between my code and David's is:  
1. the mean and standard deviation calculated in my code is using `n_patients` and `variance` while David's code is using `mean = true_theta * info_increment` and standard deviation is `sd = sqrt(info_increment)`.  
2. the boundaries and simulation is occuring on the score scale for David's code while it's occuring on the standard Z scale in mine.


Results on the score scale for my code are (and were) correct.
